---
title: "Coursera Machine Learning Assignment"
author: "KVdG"
date: "May 27 2018"
output:
  html_document:
    toc: yes
    toc_depth: 5
  pdf_document:
    toc: yes
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Init, include=FALSE}
library(caret)
library(dplyr)
library(corrplot)
library(rpart.plot)
```


# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 


# Some information about the data set

(The text below is copied from the HAR website.)

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Read more: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har#ixzz5GisaTutu

# Data Loading and Preparation

```{r DataLoad, cache=TRUE}

f1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
f2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
pml_training.csv <- read.csv(file = f1, header = TRUE )
pml_testing.csv <- read.csv(file = f2, header = TRUE )
rm(f1, f2)

```

```{r DataPrep, cache=TRUE}

## remove columns with subject variables (the first 7)
training <- pml_training.csv[ , -c(1:7) ]
testing <- pml_testing.csv[ , -c(1:7) ]

## remove columns with NAs
NA_training <- which(colSums(is.na(training)) == 0)
NA_testing <-  which(colSums(is.na(testing)) == 0)

# There are more columns with NAs in the test set, hence we take this list
# to clean both train and test set
training <- training[, NA_testing]
testing <- testing[, NA_testing]

```

<!-- ## A quick visualisation -->

<!-- It was impossible to include a detailed visualisation (too big for github?) -->

<!-- For people unfamiliar with this visualisation: the higher the correlation, the darker the color (red for negative, blue for positive). Hence, the lower correlations are less visible. -->


```{r QuickViz, eval=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
 
# names(grep("factor", sapply(training, class), value=TRUE))

### cor graph
cex.before <- par("cex")
par(cex = 0.3)
corrplot(cor( training  %>% select(-classe) ), title="\nCorrelations", 
             type="lower", method="number")
par(cex = cex.before)

```

# Model Building

## Trees

```{r model_tree, cache=TRUE}

fit_tree <- train(classe ~ . , method = "rpart", data = training)
rpart.plot(fit_tree$finalModel)
confusionMatrix(fit_tree)
rpart.plot(fit_tree$finalModel)
```





## RandomForest with parallel processing
### Prepare for parallel processing
```{r PrepParallel, warning=FALSE}

library(doParallel)
library(parallel)
detectCores(); detectCores(logical = FALSE)

```


### Build RandomForest Model
```{r RandomForest, cache=TRUE}

cluster <- makeCluster(detectCores()-1) # convention to leave 1 core for OS
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

fit_rf <- train(classe ~ . , method = "rf", data = training, trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()

confusionMatrix(fit_rf$finalModel$predicted, training$classe)

```

## Performance of models

1. The accuracy of the Tree model was 0.498
2. The accuracy of the Random Forest Model was 0.996
We continue with the latter model to do predictions

**The out of sample error for the Random Forest Model is: 0.4 percent**

# Predict with the best model: Random Forest Based
```{r PredictRF}

# print(fit_rf$finalModel)

pred_rf <- predict(fit_rf, newdata = testing)
print(pred_rf)

```

# Conclusion

Several models were tested, two were selected for reporting as they provide insight. The tree model has limited accuracy, while the random forest model is very good in predicting. 